Corotating Clumps Around Adolescent Low-Mass Stars: Four Years of Complex Quasiperiodic Variables from TESS

### 2023-08-11 paper/ms.tex
update ms

% 250 word max!
\begin{abstract}
Complex quasiperiodic variables (CQVs) are low-mass pre-main-sequence
stars with nearly periodic optical modulation.  The modulation is
likely induced by dust or gas clumps in orbit at the Keplerian
corotation radius.  Here, we report new CQVs discovered in TESS
short-cadence data collected between July~2018 and Sep~2022.  Our
search of \nstarssearched\ K and M dwarfs with $T$$<$16 and
$d$$<$150\,pc yielded \ngoods\ high-quality CQVs.  Most of these
discoveries are new, and they include the brightest
($T$$\approx$9.5), closest ($d$$\approx$20\,pc), most massive
($\approx$0.6-0.8\,$M_\odot$), and the oldest ($\approx$200\,Myr)
examples of this class of object currently known.  A few objects are
outliers: LP 12-502 for instance showed a ``dip complex'' with a
period and duration that were fixed over 1,500 cycles, but in
detail, this system exhibited between four and eight local minima
per cycle, and at times also displayed multiple periods
simultaneously.  We demonstrate that transient corotating material
is the most viable explanation for this class of object, and present
evidence supporting the hypothesis that this material is sculpted by
stellar magnetic fields dominated by high order multipoles.  We
expect that our sample will facilitate future efforts aimed at
connecting CQVs to the broader contexts of star, disk, and perhaps
even exoplanet evolution. 	
\end{abstract}

@@

\begin{figure*}[!t]
	\begin{center}
		\centering
		\includegraphics[width=0.9\textwidth]{f7.pdf}
		\vspace{-0.1cm}
		\caption{
      {\bf Properties of CQVs identified by our search}.
      CQVs are pre-main-sequence K and M dwarfs, younger than
      $\approx$200 Myr, with rotation periods faster than around half
      of Pleiades-age M dwarfs.
      The \ngoods\ bonafide CQVs in Table~\ref{tab:thetable} are the
      dark blue circles; \nmaybes\ ambiguous CQV candidates are light
      blue circles.  The top panels show the \nstarssearched\ target
      stars with 120-second cadence TESS data as the shaded gray
      background; darker regions correspond to a larger relative
      number of searched stars.  The lower-left panel compares the
      rotation--color distribution of CQVs against the rotation
      periods of K and M dwarfs in the Pleiades from
      \citet{2016AJ....152..114R}.  The lower-middle panel plots the
      derived corotation radii $R_{\rm cr} = (GM/\Omega^2)^{1/3}$ in
      units of stellar radii against the measured CQV periods, in
      units of hours.  Ages in the final panel are known from cluster
      membership.
		}
		\label{fig:catalogscatter}
	\end{center}
\end{figure*}

@@

TIC~425937691 is a suprising case: it shows two seemingly independent
CQV signals on what appears to be the same star.  The two signals have
periods of $\approx$4.816\,hr and $\approx$3.218\,hr, respectively.
The former is the larger amplitude signal visible in
Figures~\ref{fig:cqvs} and~\ref{fig:evoln}, corresponding to signal
with peak-to-peak amplitude of $\approx$8\%.  The latter signal is
only visible after subtracting the primary period, and it has a
sinusoidal envelope with a peak-to-peak amplitude of 2.5\%, and with
three sharp dips super-posed over each cycle.  The Gaia DR3 RUWE for
TIC~425937691 is 2.9.

@@

@@

\begin{enumerate}[leftmargin=*]
	%
	\item CQVs are quasiperiodic.  The mean periods remain fixed over the
		$>$1{,}000-cycle baseline of available observations; but the light curve shapes
		always evolve (Figure~\ref{fig:evoln}).
  %
  \item The same CQV can show multiple periods simultaneously.
     LP 12-502, for instance, showed dips with four
    distinct periods within $\pm 0.3\%$ of its fundamental period,
    sometimes simultaneously, and each lasting for up to 50 cycles
    (Figure~\ref{fig:lpriver0}).
  %
    \item CQVs evolve over timescales that are both secular
    ($>$100\,cycles) and impulsive ($<$1 cycle).  
    Dip growth seems to happen over durations of at least ten cycles, and slow dip decay can also occur.
    ``State-switches''
    correspond to dips to collapsing instantaneously, and are almost always
    linked with observed optical flares.
    This suggests that the occulting material is responding to sudden changes
    in the magnetic field.
  %
    \item The rate of dip evolution can be used to place a model-dependent
    constraint on how much material is either being accreted or ejected
    during the state changes (Section~\ref{subsec:massflux}).
    Order of magnitude estimates require at least an asteroid belt's worth
    of dust ($10^{-4}$\,$M_\oplus$) over $10^8$ years, or 100 times more 
    material if the occulting matter is gas.
  %
  \item The CQV phenomenon persists for $\gtrsim$150\,Myr, based on
    the existence of multiple CQVs in AB Dor, the Pleiades, and
    Psc-Eri (Section~\ref{subsec:starprops}).  It may even extend to
    200\,Myr, based on the one CQV we found in the Carina Near
    moving group (TIC~294328887; $\approx$200\,Myr).  The lack of
    detected CQVs in the Hyades and Praesepe suggests that the
    lifetime of the phenomenon is likely limited to the first few
    hundred million years.
	%
    \item The duty cycle for CQVs seems to be $\approx$$75$\%, based on
    the fraction of bonafide CQVs that either turned on or turned off during
    TESS re-observations, two years after the initial observation.
  %
    \item While the majority of detected CQVs have masses of
    0.1-0.5\,$M_\odot$, the upper mass limit for the CQV phenomenon
    might extend into the K dwarf regime ($<$$0.8$\,$M_\odot$).
    TIC~405754448 and TIC~405910546, with masses of $\approx$0.8 and
    $\approx$0.6\,$M_\odot$, are the two massive outliers
    (Section~\ref{subsec:extreme}), and we encourage additional
    scrutiny of these objects in future work.
  %
    \item Surprising analogs to the M  and K dwarf CQVs exist in
    magnetic B stars, but only those with multipolar field topologies.
    Since most magnetic B stars have dipolar magnetic fields,
    this suggests that the CQV dips and warps are similarly being sculpted
    by the stellar magnetic fields, and that the magnetic
    fields themselves are probably also multipolar.
  %
    \item The closest CQVs to the Sun are at distances of 15-20\,pc, and
    the brightest have $V$$\approx$12 ($J$$\approx$7.5).  This
    conclusion is an extrapolation from the CQV sample derived in this work
    being $\gtrsim$80\% complete within 30\,pc.  Expanding our analysis of
    the TESS data to the full frame images would yield a truly
    volume-limited selection function, and would expand the CQV census
    by about a factor of two within 50\,pc, and by a factor of ten
    within 100\,pc.
\end{enumerate}

>>>>

% 250 word max!
\begin{abstract}
Complex quasiperiodic variables (CQVs) are low-mass pre-main-sequence
stars with nearly periodic optical modulation.  The modulation is
likely induced by dust or gas clumps in orbit at the Keplerian
corotation radius.  Here, we report new CQVs discovered in TESS
short-cadence data collected between July~2018 and Sep~2022.  Our
search of \nstarssearched\ K and M dwarfs with $T$$<$16 and
$d$$<$150\,pc yielded \ngoods\ high-quality CQVs.  Most of these
discoveries are new, and they include the brightest
($T$$\approx$9.5), closest ($d$$\approx$20\,pc), and oldest
($\approx$200\,Myr) examples of this class of object currently
known.  A few objects are outliers: LP 12-502 for instance showed a
``dip complex'' with a period and duration that were fixed over
1,500 cycles, but in detail, this system exhibited between four and
eight local minima per cycle, and at times also displayed multiple
periods simultaneously.  We demonstrate that transient corotating
material is the most viable explanation for this class of object,
and present evidence supporting the hypothesis that this material is
sculpted by stellar magnetic fields dominated by high order
multipoles.  We expect that our sample will facilitate future
efforts aimed at connecting CQVs to the broader contexts of star,
disk, and perhaps even exoplanet evolution. 	
\end{abstract}

@@

\begin{figure*}[!t]
	\begin{center}
		\centering
		\includegraphics[width=0.9\textwidth]{f7.pdf}
		\vspace{-0.1cm}
		\caption{
      {\bf Properties of CQVs identified by our search}.
      CQVs are mostly pre-main-sequence M dwarfs, younger than
      $\approx$150 Myr, with rotation periods faster than $\approx$1\,day.
      The \ngoods\ bonafide CQVs in Table~\ref{tab:thetable} are the
      dark blue circles; \nmaybes\ ambiguous CQV candidates are light
      blue circles.  The top panels show the \nstarssearched\ target
      stars with 120-second cadence TESS data as the shaded gray
      background; darker regions correspond to a larger relative
      number of searched stars.  The lower-left panel compares the
      rotation--color distribution of CQVs against the rotation
      periods of K and M dwarfs in the Pleiades from
      \citet{2016AJ....152..114R}.  The lower-middle panel plots the
      derived corotation radii $R_{\rm cr} = (GM/\Omega^2)^{1/3}$ in
      units of stellar radii against the measured CQV periods, in
      units of hours.  Ages in the final panel are known from cluster
      membership.
		}
		\label{fig:catalogscatter}
	\end{center}
\end{figure*}

@@

TIC~425937691 is a surprising case: it shows two seemingly independent
CQV signals on what appears to be the same star.  The two signals have
periods of $\approx$4.82\,hr and $\approx$3.22\,hr, respectively.
The former is the larger amplitude signal visible in
Figures~\ref{fig:cqvs} and~\ref{fig:evoln}, corresponding to signal
with peak-to-peak amplitude of $\approx$8\%.  The latter signal is
only visible after subtracting the primary period, and it has a
sinusoidal envelope with a peak-to-peak amplitude of 2.5\%, and with
three sharp dips super-posed over each cycle.  The Gaia DR3 RUWE for
TIC~425937691 is 2.9.

@@

\subsection{TIC~435903839}
%TODO: move these to their own ``debunked'' table, I think...
%TODO: along with the IR excess ones
TIC~435903839, which we originally classified as an ``ambiguous'' CQV, is an apparently
similar case.
We initially labeled this source as having a 10.82\,hr period, because this
period minimized the dispersion in the phase-folded light curve.
However closer inspection reveals that this source is a photometric
blend of two ordinary rotating stars with $P_0$=3.60\,hr, and $P_1$=5.41\,hr.
This is a novel false positive scenario for CQVs:
two rapidly-rotating stars with near-integer ratios of rotation periods.
The beat between the two rotation signals produces the spurious
CQV signal.
Such false positives can be excluded through careful accounting of
all peaks in a periodogram.
For instance, TIC~435903839 shows a peak at 16.27\,hr, which is not
an integer multiple of the nominal ``10.82\,hr'' period.

@@

\begin{enumerate}[leftmargin=*]
	%
	\item CQVs are quasiperiodic.  The mean periods remain fixed over the
		$>$1{,}000-cycle baseline of available observations; but the light curve shapes
		always evolve (Figure~\ref{fig:evoln}).
  %
  \item The same CQV can show multiple periods simultaneously.
     LP 12-502, for instance, showed dips with four
    distinct periods within $\pm 0.3\%$ of its fundamental period,
    sometimes simultaneously, and each lasting for up to 50 cycles
    (Figure~\ref{fig:lpriver0}).
  %
    \item CQVs evolve over timescales that are both secular
    ($>$100\,cycles) and impulsive ($<$1 cycle).  
    Dip growth seems to happen over durations of at least ten cycles, and slow dip decay can also occur.
    ``State-switches''
    correspond to dips to collapsing instantaneously, and are almost always
    linked with observed optical flares.
    This suggests that the occulting material is responding to sudden changes
    in the magnetic field.
  %
    \item The rate of dip evolution can be used to place a model-dependent
    constraint on how much material is either being accreted or ejected
    during the state changes (Section~\ref{subsec:massflux}).
    Order of magnitude estimates require at least an asteroid belt's worth
    of dust ($10^{-4}$\,$M_\oplus$) over $10^8$ years, or 100 times more 
    material if the occulting matter is gas.
  %
  \item The CQV phenomenon persists for $\gtrsim$150\,Myr, based on
    the existence of multiple CQVs in AB Dor, the Pleiades, and
    Psc-Eri (Section~\ref{subsec:starprops}).  It may even extend to
    200\,Myr, based on the one CQV we found in the Carina Near
    moving group (TIC~294328887; $\approx$200\,Myr).  The lack of
    detected CQVs in the Hyades and Praesepe suggests that the
    lifetime of the phenomenon is likely limited to the first few
    hundred million years.
	%
    \item The duty cycle for CQVs seems to be $\approx$$75$\%, based on
    the fraction of bonafide CQVs that either turned on or turned off during
    TESS re-observations, two years after the initial observation.
  %
    \item While the majority of detected CQVs have masses of
    0.1-0.5\,$M_\odot$, the upper mass limit for the CQV phenomenon
    might extend into the K dwarf regime ($<$$0.8$\,$M_\odot$).
    TIC~405754448 and TIC~405910546, with masses of $\approx$0.8 and
    $\approx$0.6\,$M_\odot$, are the two massive outliers
    (Section~\ref{subsec:extreme}), and we encourage additional
    scrutiny of these objects in future work.
  %
    \item Surprising analogs to the CQVs exist in magnetic B stars,
    but only those with multipolar field topologies.  Since most
    magnetic B stars have dipolar magnetic fields, this suggests
    that the CQV dips and warps are similarly being sculpted by the
    stellar magnetic fields, and that the magnetic fields themselves
    are probably also multipolar.
  %
    \item The closest CQVs to the Sun are at distances of 15-20\,pc, and
    the brightest have $V$$\approx$12 ($J$$\approx$7.5).  This
    conclusion is an extrapolation from the CQV sample derived in this work
    being $\gtrsim$80\% complete within 30\,pc.  Expanding our analysis of
    the TESS data to the full frame images would yield a truly
    volume-limited selection function, and would expand the CQV census
    by about a factor of two within 50\,pc, and by a factor of ten
    within 100\,pc.
\end{enumerate}
